---
title: "Lab 04 - Data Visualization"
output:
  html_document:
    html_preview: false
link-citations: yes
---

```{r setup, message=FALSE, echo = FALSE, warning=FALSE, eval = TRUE}
library(data.table)
library(leaflet)
library(tidyverse)
```

# Learning Goals

-   Read in and prepare the meteorological dataset
-   Create several graphs with different `geoms()` in `ggplot2`
-   Create a facet graph
-   Conduct some customizations of the graphs
-   Create a more detailed map using `leaflet()`

# Lab Description

We will again work with the meteorological data presented in lecture.

**The objective of the lab is to examine the association between weekly average dew point temperature and wind speed in four regions of the US and by elevation.**

# Steps

### 1. Read in the data

First download and then read in with data.table:fread()

This is the same met_data as we used in lab 03, so you can choose to move that file to your current project directory instead of re-downloading.

```{r checking-file, include=FALSE}
fn <- "https://raw.githubusercontent.com/JSC370/jsc370-2022/main/labs/lab03/met_all.gz"
if (!file.exists("met_all.gz"))
  download.file(fn, destfile = "met_all.gz")
met <- data.table::fread("met_all.gz")
```

```{r, echo=T, message=FALSE, eval=FALSE}
download.file(
  "https://raw.githubusercontent.com/JSC370/jsc370-2022/main/labs/lab03/met_all.gz",
  destfile = "met_all.gz",
  method   = "curl",
  timeout  = 60
  )

met <- data.table::fread("met_all.gz")
```

### 2. Prepare the data

-   Remove temperatures less than -17C
-   Make sure there are no missing data in the key variables coded as 9999, 999, etc
-   Generate a date variable using the functions `as.Date()` (hint: You will need the following to create a date `paste(year, month, day, sep = "-")`).
-   Using the `data.table::week` function, keep the observations of the first week of the month.
-   Compute the mean by station of the variables `temp`, `rh`, `wind.sp`, `vis.dist`, `dew.point`, `lat`,`lon`, and `elev`.
-   Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
-   Create a categorical variable for elevation (breakpoint 252 m)

```{r, eval = TRUE, echo = FALSE}


```

### 3. Use `geom_violin` to examine the wind speed and dew point temperature by region

You saw how to use `geom_boxplot` in class. Try using `geom_violin` instead (take a look at the help). (hint: You will need to set the `x` aesthetic to 1)

-   Use facets
-   Make sure to deal with `NA` category
-   Describe what you observe in the graph

```{r, eval = TRUE, echo = FALSE}

```

### 4. Use `geom_jitter` with `stat_smooth` to examine the association between dew point temperature and wind speed by region

-   Color points by region
-   Make sure to deal with `NA` category
-   Fit a linear regression line by region
-   Describe what you observe in the graph

```{r, eval = TRUE, echo = FALSE}

  
```

### 5. Use `geom_bar` to create barplots of the weather stations by elevation category coloured by region

-   Bars by elevation category using `position="dodge"`
-   Change colors from the default. Color by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
-   Create nice labels on axes and add a title
-   Try a different theme
-   Describe what you observe in the graph
-   Make sure to deal with `NA`

```{r warning=FALSE, message=FALSE, eval = TRUE, echo = FALSE}


```

### 6. Use `stat_summary` to examine mean dew point and wind speed by region with standard deviation error bars

-   Make sure to remove `NA`
-   Use fun.data="mean_sdl" in `stat_summary`
-   Add another layer of `stats_summary` but change the geom to `"errorbar"` (see the help).
-   Describe the graph and what you observe

```{r, eval = TRUE, echo = TRUE}

```

-   Dew point temperature is...
-   Wind speed is...

### 7. Make a map showing the spatial trend in relative h in the US

-   Make sure to remove `NA`
-   Use leaflet()
-   Make a colour palette with custom colours
-   Use `addMarkers` to include the top 10 places in relative h (hint: this will be useful `rank(-rh) <= 10`)
-   Add a legend

```{r, eval = TRUE, echo = FALSE}

```

-   Describe trend in RH across the US

### 8. Use a ggplot extension

-   Pick and extension (except cowplot) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the met data (or met_avg)
-   Might want to try examples that come with the extension first (e.g. ggtech, gganimate, ggforce)
