---
title: "Lab 11 - Interactive Visualization"
output: html_document
---


# Learning Goals

- Read in and process the COVID dataset from the New York Times GitHub repository
- Create interactive graphs of different types using `plot_ly()` and `ggplotly()` functions
- Customize the hoverinfo and other plot features
- Create a Choropleth map using `plot_geo()`

# Lab Description

We will work with COVID data downloaded from the New York Times. The dataset consists of COVID-19 cases and deaths in each US state during the course of the COVID epidemic.

**The objective of this lab is to explore relationships between cases, deaths, and population sizes of US states, and plot data to demonstrate this**

# Steps

## I. Reading and processing the New York Times (NYT) state-level COVID-19 data

### 0. Install and load libraries

```{r message=FALSE, echo=FALSE, warning=FALSE}

library(data.table)
library(tidyverse)
library(plotly)
library(knitr)
library(widgetframe)
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache=FALSE,
  include=TRUE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```
### 1. Read in the data

- Read in the COVID data with data.table:fread() from the NYT GitHub repository: "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
- Read in the state population data with data.table:fread() from the repository: "https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv""
- Merge datasets

```{r}

cv_states_readin <- as.data.frame(data.table::fread("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv") )


state_pops <- as.data.frame(data.table::fread("https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv"))

state_pops$abb <- state_pops$state
state_pops$state <- state_pops$state_name
state_pops$state_name <- NULL

cv_states <- merge(cv_states_readin, state_pops, by="state")
```

### 2. Look at the data

- Inspect the dimensions, `head`, and `tail` of the data
- Inspect the structure of each variables. Are they in the correct format?

```{r}
dim(cv_states)
head(cv_states)
tail(cv_states)
str(cv_states)
```

### 3. Format the data

- Make date into a date variable
- Make state into a factor variable
- Order the data first by state, second by date
- Confirm the variables are now correctly formatted
- Inspect the range values for each variable. What is the date range? The range of cases and deaths?

```{r}

cv_states$date <- as.Date(cv_states$date, format="%Y-%m-%d")


state_list <- unique(cv_states$state)
cv_states$state <- factor(cv_states$state, levels = state_list)
abb_list <- unique(cv_states$abb)
cv_states$abb <- factor(cv_states$abb, levels = abb_list)


cv_states = cv_states[order(cv_states$state, cv_states$date),]


str(cv_states)
head(cv_states)
tail(cv_states)


head(cv_states)
summary(cv_states)
min(cv_states$date)
max(cv_states$date)
```

### 4. Add `new_cases` and `new_deaths` and correct outliers

- Add variables for new cases, `new_cases`, and new deaths, `new_deaths`: 
  - Hint: You can set `new_cases` equal to the difference between cases on date i and date i-1, starting on date i=2
  
- Filter to dates after July 1, 2021


```{r}

for (i in 1:length(state_list)) {
  cv_subset = subset(cv_states, state == state_list[i])
  cv_subset = cv_subset[order(cv_subset$date),]

  # add starting level for new cases and deaths
  cv_subset$new_cases = cv_subset$cases[1]
  cv_subset$new_deaths = cv_subset$deaths[1]


  for (j in 2:nrow(cv_subset)) {
    cv_subset$new_cases[j] = cv_subset$cases[j] - cv_subset$cases[j-1] 
    cv_subset$new_deaths[j] = cv_subset$deaths[j] - cv_subset$deaths[j-1]
  }
  
  cv_states$new_cases[cv_states$state==state_list[i]] = cv_subset$new_cases
  cv_states$new_deaths[cv_states$state==state_list[i]] = cv_subset$new_deaths
}

# Focus on recent dates
cv_states <- cv_states %>% dplyr::filter(date >= "2021-07-01")
```

- Use `ggplotly` for EDA: See if there are outliers or values that don't make sense for `new_cases` and `new_deaths`. Which states and which dates have strange values?

```{r}

p1 <- ggplot(cv_states, aes(x = date, y = new_cases, color = state)) +
  geom_line()

ggplotly(p1)

```
There were -4678 recorded new cases on January 29, 2022 in Colorado, and -4397 recorded new cases in Pennsylvania on February 8, 2022. Neither of these values seem realistic, as we cannot have negative amounts of new cases.


```{r}

p2 <- ggplot(cv_states, aes(x = date, y = new_deaths, color = state)) +
  geom_line()

ggplotly(p2)

```

There are a few instances in which a negative amount of people were recorded as dying on a particular day, such as -3770 deaths in Massachusetts on March 14, 2022, or -357 deaths in California on August 11, 2021. 

Additionally, there are other outliers in the data where some states reported incredibly high numbers on a particular day compared to other days, such as the single-day spikes for Missouri and Tennessee.


- Correct outliers: Set negative values for `new_cases` or `new_deaths` to 0

- Inspect data again interactively

```{r}

cv_states$new_cases[cv_states$new_cases < 0] = 0
cv_states$new_deaths[cv_states$new_deaths < 0] = 0

p3 <- ggplot(cv_states, aes(x = date, y = new_deaths, color = state)) +
  geom_line() + geom_point(size = 0.5, alpha = 0.5)

ggplotly(p3)


```

### 5. Add additional variables

- Add population-normalized (by 100,000) variables for each variable type (rounded to 1 decimal place). Make sure the variables you calculate are in the correct format (`numeric`). You can use the following variable names:
  - `per100k` = cases per 100,000 population
  - `newper100k`= new cases per 100,000
  - `deathsper100k` = deaths per 100,000
  - `newdeathsper100k` = new deaths per 100,000

- Add a "naive CFR" variable representing `deaths / cases` on each date for each state

- Create a dataframe representing values on the most recent date, `cv_states_today`

```{r}

cv_states$per100k =  as.numeric(format(round(cv_states$cases/(cv_states$population/100000),1),nsmall=1))
cv_states$newper100k =  as.numeric(format(round(cv_states$new_cases/(cv_states$population/100000),1),nsmall=1))
cv_states$deathsper100k =  as.numeric(format(round(cv_states$deaths/(cv_states$population/100000),1),nsmall=1))
cv_states$newdeathsper100k =  as.numeric(format(round(cv_states$new_deaths/(cv_states$population/100000),1),nsmall=1))

cv_states = cv_states %>% mutate(naive_CFR = round((deaths*100/cases),2))

max_date <- max(cv_states$date)
cv_states_today = cv_states %>% filter(date==as.Date(max_date))

```

## II. Scatterplots

### 6. Explore scatterplots using `plot_ly()`

- Create a scatterplot using `plot_ly()` representing `pop_density` vs. various variables (e.g. `cases`, `per100k`, `deaths`, `deathsper100k`) for each state on most recent date (`cv_states_today`)
  - Color points by state and size points by state population
  - Use hover to identify any outliers. 


```{r}
cv_states_today %>% 
  plot_ly(x = ~pop_density, y = ~cases, type = 'scatter', mode = 'markers',
          color = ~state, size = ~population, sizes = c(5, 70), 
          marker = list(sizemode = 'diameter', opacity = 0.5))
```

Washington D.C. is a massive outlier compared to the 50 states, so we will remove it and plot the remaining areas.

- Remove those outliers and replot.

```{r}
cv_states_today %>% 
  filter(state != "District of Columbia") %>%
  plot_ly(x = ~pop_density, y = ~cases, type = 'scatter', mode = 'markers',
          color = ~state, size = ~population, sizes = c(5, 70), 
          marker = list(sizemode = 'diameter', opacity = 0.5))
```

- Choose one plot. For this plot:
- Add hoverinfo specifying the state name, cases per 100k, and deaths per 100k, similarly to how we did this in the lecture notes
- Add layout information to title the chart and the axes
- Enable `hovermode = "compare"`
  
```{r}
cv_states_today %>% 
  filter(state != "District of Columbia") %>%
  plot_ly(x = ~pop_density, y = ~cases, type = 'scatter', mode = 'markers',
          color = ~state, size = ~population, sizes = c(5, 70), 
          marker = list(sizemode = 'diameter', opacity = 0.5),
          hover_info = "text", 
          text = ~paste0("State: ", state,
                         "<br>Cases per 100k: ", per100k,
                         "<br>Deaths per 100k: ", deathsper100k)) %>%
  layout(title = "Population-normalized cases per 100k",
         yaxis = list(title = "Cases per 100k"),
         xaxis = list(title = "Population Density"),
         hovermode = "compare")
```

 
  
### 7. Explore scatterplot trend interactively using `ggplotly()`

- For `pop_density` vs. `newdeathsper100k` create a chart with the same variables using `ggplotly()`
- Explore the pattern between $x$ and $y$
- Explain what you see. Do you think `pop_density` correlates with `newdeathsper100k`?

```{r}
p4 <- cv_states_today %>%
  filter(state != "District of Columbia") %>%
  ggplot(aes(x = pop_density, y = deathsper100k, colour = state, size = population)) + 
  geom_point()
  
ggplotly(p4)
```


### 8. Multiple line chart

- Create a line chart of the `naive_CFR` for all states over time using `plot_ly()`
  - Use the zoom and pan tools to inspect the `naive_CFR` for the states that had an increase in September. How have they changed over time?
- Create one more line chart, for Florida only, which shows `new_cases` and `new_deaths` together in one plot. Hint: use `add_layer()`
  - Use hoverinfo to "eyeball" the approximate peak of deaths and peak of cases. What is the time delay between the peak of cases and the peak of deaths?
  

```{r}
cv_states %>%
  plot_ly(x = ~date, y = ~naive_CFR, color = ~state, type = "scatter", mode = "lines")
```

```{r}
cv_states %>% 
  filter(state == "Florida") %>%
  plot_ly(x = ~date, y = ~new_cases, type = "scatter", mode = "lines") %>%
  add_lines(x = ~date, y = ~new_deaths)
```


### 9. Heatmaps

Create a heatmap to visualize `new_cases` for each state on each date greater than July 1st, 2021
- Start by mapping selected features in the dataframe into a matrix using the **tidyr** package function `pivot_wider()`, naming the rows and columns, as done in the lecture notes
- Use `plot_ly()` to create a heatmap out of this matrix. Which states stand out?


```{r}

cv_states_mat <- cv_states %>%
  select(state, date, new_cases) %>%
  filter(date > "2021-07-01")

cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state,
                                            values_from = new_cases))

cv_states_mat2 <- cv_states_mat2 %>%
  column_to_rownames('date') %>%
  as.matrix()

plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2), 
        z = ~cv_states_mat2, type = 'heatmap')

```

- Create a second heatmap in which the pattern of `new_cases` for each state over time becomes more clear by filtering to only look at dates every two weeks (check out the omicron wave).

```{r}

filter_dates <- seq(as.Date("2021-11-01"), as.Date("2022-03-22"), by="2 weeks")

cv_states_mat <- cv_states %>% select(state, date, new_cases) %>% filter( date %in% filter_dates )
cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state, values_from = new_cases))
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date <- NULL
cv_states_mat2 <- as.matrix(cv_states_mat2)

#create heatmap
```

### 10. Map

- Create a map to visualize the `naive_CFR` by state on Devember 15, 2021
- Compare with a map visualizing the `naive_CFR` by state on most recent date


```{r}

pick.date = "2021-12-15"

# Extract the data for each state by its abbreviation
cv_per100 <- cv_states %>% filter(date==pick.date) %>% select(state, abb, newper100k, cases, deaths) # select data
cv_per100$state_name <- cv_per100$state
cv_per100$state <- cv_per100$abb
cv_per100$abb <- NULL

# Create hover text
cv_per100$hover <- with(cv_per100, paste(state_name, '<br>', "Cases per 100k: ", newper100k, '<br>', "Cases: ", cases, '<br>', "Deaths: ", deaths))

# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('steelblue')
)

# Make sure both maps are on the same color scale
shadeLimit <- 125

# Create the map

```

```{r}
# Map for today's date

# Extract the data for each state by its abbreviation
cv_per100 <- cv_states_today %>%  select(state, abb, newper100k, cases, deaths) # select data
cv_per100$state_name <- cv_per100$state
cv_per100$state <- cv_per100$abb
cv_per100$abb <- NULL

# Create hover text
cv_per100$hover <- with(cv_per100, paste(state_name, '<br>', "Cases per 100k: ", newper100k, '<br>', "Cases: ", cases, '<br>', "Deaths: ", deaths))

# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('steelblue')
)

# Create the map

```

# Deliverables 

Lab 10b questions 1-2, lab 11 questions 0-9. Upload html or pdf for both lab Rmd's to quercus. 
