---
title: "Lab 05 - Data Wrangling"
output:
  html_document:
    html_preview: false
link-citations: yes
---

# Learning goals

-   Use the `merge()` function to join two datasets.
-   Deal with missings and impute data.
-   Identify relevant observations using `quantile()`.
-   Practice your GitHub skills.

# Lab description

For this lab we will be dealing with the meteorological dataset `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup a Git project and the GitHub repository

1.  Go to wherever you are planning to store the data on your computer, and create a folder for this project, for example, "JSC370-labs"

2.  In that folder, save [this lab template](https://github.com/JSC370/jsc370-2022/blob/main/labs/lab05/lab05-wrangling-gam.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3.  Go to your GitHub account and create a new repository of the same name that your local folder has, e.g., "JSC370-labs".

4.  Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5.  Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Most of the steps can be done using command line:

``` {.sh}
# Step 1
cd ~/Documents
mkdir JSC370-labs
cd JSC370-labs

# Step 2
wget https://github.com/JSC370/jsc370-2022/blob/main/labs/lab05/lab05-wrangling-gam.Rmd
mv lab05-wrangling-gam.Rmd README.md

# Step 3
# Happens on github

# Step 4
git init
git add README.Rmd
git commit -m "First commit"

# Step 5
git remote add origin git@github.com:[username]/JSC370-labs
git push -u origin master
```

You can also complete the steps in R (replace with your paths/username when needed)

``` {.r}
# Step 1
setwd("~/Documents")
dir.create("JSC370-labs")
setwd("JSC370-labs")

# Step 2
download.file(
  "https://github.com/JSC370/jsc370-2022/blob/main/labs/lab05/lab05-wrangling-gam.Rmd",
  destfile = "README.Rmd"
  )

# Step 3: Happens on Github

# Step 4
system("git init && git add README.Rmd")
system('git commit -m "First commit"')

# Step 5
system("git remote add origin git@github.com:[username]/JSC370-labs")
system("git push -u origin master")
```

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1.  Load the `data.table` (and the `dtplyr` and `dplyr` packages if you plan to work with those).

```{r setup}
library(dtplyr)
library(dplyr)
library(data.table)
```

2.  Load the met data from <https://github.com/JSC370/jsc370-2022/blob/main/labs/lab03/met_all.gz>, and also the station data. For the latter, you can use the code we used during lecture to pre-process the stations data:

```{r stations-data, warning = FALSE}
# library(data.table)
# Download the data
stations <- data.table::fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]
```

3.  Merge the data as we did during the lecture.

## Question 1: Representative station for the US

Across all weather stations, what is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the `quantile()` function. Do these three coincide?

```{r}
# Read in data
met <- data.table::fread("met_all.gz")

# Merge data
weather_stations <- merge(x = met, y = stations, 
                          all.x = TRUE, all.y = FALSE,
                          by.x = "USAFID", by.y = "USAF")
```

```{r}
# Compute means
station_avgs <- weather_stations[, .(temp = mean(temp, na.rm = TRUE),
                                     wind.sp = mean(wind.sp, na.rm = TRUE),
                                     atm.press = mean(atm.press, na.rm = TRUE)),
                                 by = .(USAFID, STATE)]
```

```{r}
# Compute medians
station_medians <- station_avgs[, .(temp_50 = quantile(temp, probs = 0.50, na.rm = TRUE),
                                    wind.sp = quantile(wind.sp, probs = 0.50, na.rm = TRUE),
                                    atm.press = quantile(atm.press, probs = 0.50, na.rm = TRUE))]
```

Knit the document, commit your changes, and save it on GitHub. Don't forget to add `README.md` to the tree, the first time you render it.

## Question 2: Representative station per state

Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

```{r}
# Finding the station closest to median temperature
station_avgs[, temp_dist := abs(temp - station_medians$temp_50)]

# Finding the station closest to median wind speed
station_avgs[, wind.sp_dist := abs(wind.sp - station_medians$wind.sp)]

# Finding the station closest to median atmospheric pressure
station_avgs[, atm.press_dist := abs(atm.press - station_medians$atm.press)]

median_temp_station <- station_avgs[order(temp_dist)][1]
median_wind_sp_station <- station_avgs[order(wind.sp_dist)][1]
median_atm_press_station <- station_avgs[order(atm.press)][1]

median_temp_station
median_wind_sp_station
median_atm_press_station
```

From the outputs above, we can see that these median stations are not the same. The weather station closest to the median temperature was in Kentucky, the station closest to the median wind speed is in Wisconsin, and the station closes to the median atmospheric pressure is in Illinois.

Knit the doc and save it on GitHub.

## Question 3: In the middle?

For each state, identify what is the station that is closest to the mid-point of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all \~100 points in the same figure, applying different colours for those identified in this question.

```{r}
# Finding midpoints per state
geographic_midpoint <- weather_stations[, .(lon_50 = quantile(lon, probs = 0.50, na.rm = TRUE),
                                            lat_50 = quantile(lat, probs = 0.50, na.rm = TRUE)),
                                        by = STATE]
midpoints <- merge(x = weather_stations, y = geographic_midpoint, by = "STATE")

# Finding Euclidean Distances
midpoints[, distance_from_mid := sqrt((lon - lon_50)^2 + (lat - lat_50)^2)]
midpoint_station <- midpoints[, .SD[which.min(distance_from_mid)], by = "STATE"]
```

```{r, warning = FALSE}
library(leaflet)
# Plotting the mid-point stations of each state on a map

leaflet() %>%
  addProviderTiles('CartoDB.Positron') %>%
  addCircles(data = midpoint_station, lat = ~lat, lng = ~lon, popup = "Midpoint of Weather Stations in this state",
             opacity = 1, fillOpacity = 1, radius = 400, color = "Blue")
```

Knit the doc and save it on GitHub.

## Question 4: Means of means

Using the `quantile()` function, generate a summary table that shows the number of states included, average temperature, wind-speed, and atmospheric pressure by the variable "average temperature level," which you'll need to create.

Start by computing the states' average temperature. Use that measurement to classify them according to the following criteria:

-   low: temp \< 20
-   Mid: temp \>= 20 and temp \< 25
-   High: temp \>= 25

```{r}
weather_stations[, mean_state_temp := mean(temp, na.rm = T), by = "STATE"]
weather_stations[, temperature_cat := fifelse(mean_state_temp < 20, "low-temp",
                                              fifelse(mean_state_temp < 25, "mid-temp",
                                                      "high-temp"))]
```

Once you are done with that, you can compute the following:

-   Number of entries (records),
-   Number of NA entries,
-   Number of stations,
-   Number of states included, and
-   Mean temperature, wind-speed, and atmospheric pressure.

For each of the temperature levels described before.

```{r}
table(weather_stations$temperature_cat, useNA = "always")

info_table <- weather_stations[, .(N_entries = .N,
                                   N_entries_na = sum(is.na(temperature_cat)),
                                   N_stations = length(unique(USAFID)),
                                   N_states = length(unique(STATE)),
                                   mean_temp = mean(temp, na.rm = TRUE),
                                   mean_wind.sp = mean(wind.sp, na.rm = TRUE),
                                   mean_atm.press = mean(atm.press, na.rm = TRUE)),
                               by = temperature_cat]

knitr::kable(info_table, caption = "Information about Weather Stations in the Continental U.S.",
             col.names = c("Temperature Category", "Number of Entries", "Number of Missing Entries",
                           "Number of Weather Stations", "Number of States", "Mean Temperature",
                           "Mean Wind Speed", "Mean Atmospheric Pressure"))

```

Knit the document, commit your changes, and push them to GitHub.

## Question 5: Advanced Regression

Let's practice running regression models with smooth functions on X. We need the `mgcv()` package and `gam()` function to do this.

-   using your data with the median values per station, examine the association between median temperature (y) and median wind speed (x). Create a scatterplot of the two variables using ggplot2. Add both a linear regression line and a smooth line.

-   fit both a linear model and a spline model (use `gam()` with a cubic regression spline on wind speed). Summarize and plot the results from the models and interpret which model is the best fit and why.

```{r}
station_avgs[, temp_50 := quantile(temp, probs = 0.50, na.rm = TRUE), by = 'STATE']
station_avgs[, wind.sp_50 := quantile(wind.sp, probs = 0.50, na.rm = TRUE), by = 'STATE']

library(ggplot2)
ggplot(station_avgs, aes(x = wind.sp_50, y = temp_50)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  geom_smooth(method = "gam", col = 2) +
  theme_bw()
```

```{r}
library(mgcv)

linear_model <- lm(temp_50 ~ wind.sp_50, data = station_avgs)
summary(linear_model)
plot(linear_model)

gam_model <- gam(temp_50 ~ s(wind.sp_50, bs = "cr", k = 20), data = station_avgs)
summary(gam_model)
plot(gam_model)
```

Since the R-squared value for the GAM was 0.67, and the R-squared value for the linear model was less than 0.02, we see that the GAM is a significantly better fit for the data than the linear model, which makes sense logically, as a GAM should be a better fit for data which is not necessarily linear.

# Deliverables

-   Answer questions 1-5

-   Paste the link to the github repo you created here

This file is part of my larger repo for the entire course. The link to my JSC370 repo on GitHub is [https://github.com/hotz1/JSC370-Repo](https://github.com/hotz1/JSC370-Repo), and the link to this particular lab is [https://github.com/hotz1/JSC370-Repo/tree/main/Labs/lab05](https://github.com/hotz1/JSC370-Repo/tree/main/Labs/lab05).

-   Submit pdf or html output to Quercus
